Prognostic and biological relevance of collagen biosynthesis pathway in prostate cancer

### 2024-01-01 report/markdown/manuscript_figures_tables.Rmd
Report update
Main figures as discussed with the study team, abbreviated Supplementary Material

---
title: "Prognostic and biologic relevance of collagen biosynthesis pathway in prostate cancer"
subtitle: "Bioinformatic methods, figures and tables"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

# Tables

```{r tab-cohorts, tab.cap = 'Characteristic of the analyzed cohorts. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Qualitative variables are presented as percentages of categories within the complete observation set.'}

flextable::flextable(paper_tbl$cohorts) %>% 
  set_widths(c(3, rep(3.5, 5))) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('PSA: prostate-specific antigen.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  theme_vanilla

```{r tab-genes, tab.cap = 'Collagen genes of interest and their classification.'}

flextable::flextable(paper_tbl$genes) %>% 
  set_widths(rep(4, 3)) %>% 
  italic(j = 1, part = 'body') %>% 
  footnote(1, 3, 
           value = as_paragraph('ECM: extracellular matrix.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

```{r fig-norm-tumor, fig.width = figur::convert(paper_fig$norm_tumor, to = 'in')$w, fig.height = figur::convert(paper_fig$norm_tumor, to = 'in')$h, fig.cap = 'Differences in tissue diffusion capacity and expression of the collagen pathway genes between the prostate cancer and benign tissue.'}

@@

__Figure \@ref(fig:fig-norm-tumor). Differences in tissue diffusion capacity and expression of the collagen pathway genes between the prostate cancer and benign tissue.__

_(A) Representative MRI axial sequences of the prostate of an 83-year-old patient with prostate cancer (left panel). The T2-weighted images (a) show a 23mm suspicious hypo-intense lesion in the left peripheral zone (red star) of the prostate, which has an early enhancement in the contrast-enhanced T1-weighted image (b). The DWI scan (c) and ADC map (d) show a strong diffusion restriction (black and white character) of the tumor area suggesting a PI-RADS 5 lesion. Statistical significance for differences in diffusion capacity in patient-matched tumor and benign tissue measured by ADC was determined by paired T test with Cohen's d effect size statistic (right panel). Single ADC values are visualized as points, grey lines connect measurements of the same donors. The number of measurement pairs, effect size and p value are displayed in the plot caption._

_(B) Differences in $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` genes related to collagen metabolism between donor-matched tumor and benign prostate tissue were investigated by paired T test with Cohen's d effect size statistic in three published cohorts. P values were corrected for multiple testing with the false discovery method. Results for gene significantly regulated in at least two cohorts are presented. Lines represent means, 2 $\times$ SEM (standard error of the mean) intervals are visualized as tinted ribbons. Gene symbols, effect sizes and p values are shown in the Y axes. Significant effects are highlighted in bold. Numbers of tissue pairs are displayed in the plot captions._

@@

```{r fig-coll-clusters, fig.width = figur::convert(paper_fig$coll_clusters, to = 'in')$w, fig.height = figur::convert(paper_fig$coll_clusters, to = 'in')$h, fig.cap = 'Collagen clusters of prostate carcinoma and their clinical characteristic.'}

paper_fig$coll_clusters$plot

@@

__Figure \@ref(fig:fig-coll-clusters). Collagen clusters of prostate carcinoma and their clinical characteristic.__

_Prostate cancers in the training TCGA cohort were clustered in respect to normalized $log_2$-transformed expression levels of the collagen pathway genes of interest by the PAM algorithm with Manhattan distance between the observations. Three clusters, collagen^low^ (low), collagen^intermediate^ (int) and collagen^high^ (hi) were defined. Cancer samples from the test cohorts (`r globals$study_labels[globals$study_labels != 'TCGA'] %>% paste(collapse = ', ')`) were assigned to the collagen clusters with an inverse distance weighted k-nearest neighbor classifier._

_(A) Mean normalized $log_2$-transformed expression levels of the collagen pathway genes in the collagen clusters and cohorts presented in a heat map. ECM: extracellular matrix._

_(B) Distribution of the collagen clusters expressed as percentage of the cohort cancer samples visualized in a stack plot. Total numbers of cancer samples are displayed in the X axis. _

_(C, D) Distribution of Gleason scores (GS, C) and pathological tumor stages (pT, D) in the collagen clusters. Statistical significance was determined by $\chi^2$ test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Gleason score and tumor stage distributions are shown in stack plots. Effect sizes and p values are displayed in the plot captions. Numbers of observations in the clusters are indicated in the X axes._

@@

```{r fig-cluster-biology, fig.width = figur::convert(paper_fig$cluster_biology, to = 'in')$w, fig.height = figur::convert(paper_fig$cluster_biology, to = 'in')$h, fig.cap = 'Infiltration, biological processes, signaling and metabolic pathways differentially regulated in the collagen clusters.'}

paper_fig$cluster_biology$plot

@@

__Figure \@ref(fig:fig-cluster-biology). Infiltration, biological processes, signaling and metabolic pathways differentially regulated in the collagen clusters.__

_(A) Counts of cancer-associated fibroblasts (CAF) and endothelial cells (EC) estimated by the MCP counter algorithm were found to differ significantly between the collagen clusters in all investigated collectives. Statistical significance was assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic. P values were corrected for multiple comparison with the false discovery rate method. Counts of fibroblasts and endothelial cells in the collagen clusters in the TCGA cohort are shown in violin plots. Red diamonds with whiskers represented medians with interquartile ranges. Single tumor samples are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of observations in the clusters are indicated in the X axes. Full analysis results are presented in Supplementary Table S4._

_(B) Gene set variation analysis (GSVA) with Reactome pathway gene signatures. Statistical significance for differences in signature single sample gene set enrichment analysis scores (ssGSEA) between the collagen clusters was determined by one-way ANOVA with $\eta^2$ effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Mean ssGSEA scores in the clusters and cohorts for signatures significantly regulated in at least four cohorts are presented in a heat map. Full analysis results are presented in Supplementary Table S6. ECM: extracellular matrix, GF: growth factor._

_(C) Modulation of KEGG-listed signaling pathways in the collagen^high^ (hi) or collagen^intermediate^ (hi) as compared with collagen^low^ tumors was predicted by the SPIA algorithm based on differential gene expression estimates. Predicted magnitude of pathway modulation (tA parameter) for pathways significantly modulated in at least four cohorts was visualized in bubble plots (tA < 0: inhibition, blue; tA > 0, activation, red). Points are labeled with their tA values. Significant effects are highlighted in bold. Full analysis results are presented in Supplementary Table S8. ECM: extracellular matrix, ACT: actin, SCLC: small cell lung carcinoma, S. aureus: Staphylococcus aureus, SLE: systemic lupus erythematosus._

_(D) Modulation of Recon2 model metabolic reactions in the collagen^high^ (hi) or collagen^intermediate^ (hi) as compared with collagen^low^ tumors was predicted by the BiGGR and biggrExtra algorithms based on differential gene expression estimates. Enrichment of significantly activated and significantly inhibited reactions in the collagen^high^ or collagen^intermediate^ clusters within the Recon model metabolic subsystems was investigated by Fisher's exact test. Enrichment p values were corrected for multiple testing with the false discovery rate method. Enrichment odds ratios for activated (red) and inhibited reaction (blue) for significantly enriched metabolic subsystems shared by at least four cohorts are presented in bubble plots. Points are labeled with their odds ration values. Full analysis results are presented in Supplementary Table S10. Val: valine, Leu: leucine, Ile: isoleucine, FAOX: fatty acid oxidation._

>>>>

---
title: "Prognostic and biological relevance of collagen biosynthesis pathway in prostate cancer"
subtitle: "Methods, figures and tables for the transcriptome part"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

# Methods

## Magnetic resonance imaging of the prostate

_Dear co-authors, it's your part as experts:)_

ADC was compared between patient-matched measurements of benign prostate and malignancy by paired T test with Cohen's d effect size statistic.

## Transcriptome analyses

The transcriptome analysis was performed with R version 4.2.3. 
Its details are presented in __Supplementary Methods__.

## Transcriptome data sets

Six publicly available transcriptome data sets of primary prostate cancer samples, 
the TCGA prostate cancer cohort [@Liu2018; @Abeshouse2015] 
(n = `r tcga$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), 
GSE54460 [@Long2014] (n = `r gse54460$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), 
GSE70768  [@Ross-Adams2015] (n = `r gse70768$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
GSE70769  [@Ross-Adams2015] (n = `r gse70769$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
GSE220095 [@Schimmelpfennig2023] (n = `r gse220095$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), 
DKFZ [@Gerhauser2018] (n = `r dkfz$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing) 
were re-analyzed (__Supplementary Table S1__). 
Normalized gene expression data and clinical information were fetched from the cBioportal repository (DKFZ, TCGA) or Gene Expression Omnibus (GEO). 
Selection criteria of the data sets were: availability of Gleason scoring, information of biochemical relapse and relapse-free survival, and availability of expression data for `r nrow(globals$genes_interest)` collagen-related genes covered by proteomic analysis as well (__Supplementary Table S2__).
Donor-matched benign and malignant tissue samples were available for the 
GSE70768 (n = `r norm_tumor$n_numbers$gse70768` tissue pairs) and TCGA cohort (n = `r norm_tumor$n_numbers$tcga` tissue pairs). 
Expression values were $log_2$-transformed prior to analyses. 
Non-immune cell content in cancer samples was estimated with the _MCP Counter_ and _xCell_ algorithms [@Aran2017; @Becht2016].
single sample gene set enrichment scores (ssGSEA) for the Reactome pathway gene signatures retrieved from the MSig database version 7.5.1 were computed with the _GSVA_ algorithm [@Hanzelmann2013].

The collagen-related genes were retrieved from the published Collagen Signature [@Kocher2021] and the Reactome collagen formation pathway (R-HSA-1474290) and restricted to genes covered by the proteomic analyses (__Supplementary Table S2__).

## Differential expression in benign and malignant prostate tissue and cancers stratified by Gleason score

Expression of the collagen-related genes of interest between patient-matched benign and malignant specimens was compared by the false discovery rate method (FDR) corrected paired two-tailed T test with Cohen's d effect size statistic. 
Differentially expressed genes were defined by pFDR < 0.05 and d $\geq$ 0.2 criteria indicative of significant, at least weak expression regulation (__Supplementary Table S3__).

Differential gene expression in cancer samples stratified by Gleason scoring (5 - 6, 7 and $\geq$ 8) was assessed by FDR-corrected one-way ANOVA with $\eta^2$ effect size statistic. 
Genes with pFDR < 0.05 and $\eta^2 \geq$ 0.02 were deemed differentially regulated (__Supplementary Table S4__).

## Transcriptomic Collagen Score and prediction of biochemical relapse-free survival

Biochemical relapse-free survival was modeled in the TCGA training cohort as a function of first- and second-order normalized, $\log_2$-transformed expression levels of the collagen-related genes of interest by Elastic Net multi-parameter Cox regression [@Simon2011; @Zou2005]. 
The Transcriptomic Collagen Score was computed as the linear predictor score of the Elastic Net Cox model with non-zero model coefficients presented in __Figure \@ref(fig:fig-coll-score)A__ and __Supplementary Table S15__.

Performance of the Transcriptomic Collagen Score at prediction of biochemical relapse-free survival in the training collective (TCGA) and validation data sets (remaining cohorts) was assessed by univariable Cox regression with Harrell's concordance index [@Harrell1996] and integrated Brier score [@Graf1999] as metrics of goodness of fit and calibration (__Supplementary Table 16__).

## Data and code availability

Publicly available data sets were analyzed. 
Formatted data sets used for analyses will be made available upon request to the corresponding author. 
The transcriptome R analysis pipeline is available from GitHub (https://github.com/PiotrTymoszuk/collagen_pca).

@@

```{r fig-norm-tumor, fig.width = figur::convert(paper_fig$norm_tumor, to = 'in')$w, fig.height = figur::convert(paper_fig$norm_tumor, to = 'in')$h, fig.cap = 'Differences in tissue diffusion capacity and expression of collagen-related genes between the prostate cancer and benign tissue. Expression of collagen-related genes in cancers stratified by Gleason scores.'}

@@

__Figure \@ref(fig:fig-norm-tumor). Differences in tissue diffusion capacity and expression of collagen-related genes between the prostate cancer and benign tissue. Expression of collagen-related genes in cancers stratified by Gleason scores.__

_(A) Representative MRI axial sequences of the prostate of an 83-year-old patient with prostate cancer (left panel). The T2-weighted images (a) show a 23mm suspicious hypo-intense lesion in the left peripheral zone (red star) of the prostate, which has an early enhancement in the contrast-enhanced T1-weighted image (b). The DWI scan (c) and ADC map (d) show a strong diffusion restriction (black and white character) of the tumor area suggesting a PI-RADS 5 lesion. Statistical significance for differences in diffusion capacity in patient-matched tumor and benign tissue measured by ADC was determined by paired T test with Cohen's d effect size statistic (right panel). Single ADC values are visualized as points, grey lines connect measurements of the same donors. The number of analyzed benign - tumor pairs, effect size and p value are displayed in the plot caption._

_(B) Differences in $log_2$-transformed expression of `r nrow(globals$genes_interest)` collagen-related genes between donor-matched tumor and benign prostate tissue samples were investigated by paired T test with Cohen's d effect size statistic in the GSE70768 (n = `r norm_tumor$n_numbers$gse70768`) and TCGA (n = `r norm_tumor$n_numbers$tcga` tissue pairs) cohorts. P values were corrected for multiple testing with the false discovery rate method. Mean normalized $log_2$-transformed expression levels in benign and cancer tissue for `r length(unique(unlist(norm_tumor$common)))` genes significantly regulated in both cohorts with at least weak effect size (pFDR < 0.05, d $\geq$ 0.2) are presented in a heat map._

_(C) Differences in $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` collagen-related genes between cancer samples stratified by Gleason scores were assessed by one-way ANOVA with $\eta^2$ effect size statistic in the GSE54460 (`r format_strata_n(gs_uni$n_numbers$gse54460)`), GSE70768 (`r format_strata_n(gs_uni$n_numbers$gse70768)`), GSE70769 (`r format_strata_n(gs_uni$n_numbers$gse70769)`), GSE220095 (`r format_strata_n(gs_uni$n_numbers$gse220095)`), DKFZ (`r format_strata_n(gs_uni$n_numbers$dkfz)`), and TCGA cohort (`r format_strata_n(gs_uni$n_numbers$tcga)`). Mean normalized $log_2$-transformed expression levels of genes significantly regulated between the Gleason score strata with at least weak effect size (pFDR < 0.05, $\eta^2 \geq$ 0.02) are presented in a heat map._

@@

```{r fig-tme, fig.width = figur::convert(paper_fig$tme, to = 'in')$w, fig.height = figur::convert(paper_fig$tme, to = 'in')$h, fig.cap = 'Expression of collagen-related genes in components of the tumor microenvironment.'}

paper_fig$tme$plot

@@

__Figure \@ref(fig:fig-tme). Expression of collagen-related genes in components of the tumor microenvironment.__

_(A) Cancer samples in six published cohorts were classified as collagen low and collagen high tumors in respect to `r nrow(globals$genes_interest)` collagen-related transcripts by semi-supervised PAM clustering (GSE54460: `r format_strata_n(ana_globals$n_numbers$gse54460)`, GSE70768: `r format_strata_n(ana_globals$n_numbers$gse70768)`, GSE70769: `r format_strata_n(ana_globals$n_numbers$gse70769)`, GSE220095: `r format_strata_n(ana_globals$n_numbers$gse220095)`, DKFZ: `r format_strata_n(ana_globals$n_numbers$dkfz)`, TCGA: `r format_strata_n(ana_globals$n_numbers$tcga)`). Counts of cancer-associated fibroblasts and endothelial cells were predicted by the MCP Counter algorithm and compared between the collagen clusters by Mann-Whitney test with r effect size statistic. Median normalized cell counts with interquartile ranges are presented as boxes, whiskers span over 150% of the interquartile ranges. Values for single cancer samples are depicted as points. All presented effects were significant (pFDR < 0.05) with at least weak (endothelial cells, r > 0.21) or moderate effect size (fibroblasts, r > 0.55)._

_(B) Placeholder for the single cell RNA seq data._

@@

```{r fig-coll-score, fig.width = figur::convert(paper_fig$coll_score, to = 'in')$w, fig.height = figur::convert(paper_fig$coll_score, to = 'in')$h, fig.cap = 'Transcriptional Collagen Score and prediction of biochemical relapse-free survival.'}

paper_fig$coll_score$plot

@@

__Figure \@ref(fig:fig-coll-score). Transcriptional Collagen Score and prediction of biochemical relapse-free survival.__

_Transcriptional Collagen Score was established in the TCGA training cohort by multi-parameter Elastic Net Cox modeling of biochemical relapse-free survival as a function of first- and second-order normalized $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` collagen-related genes._

_(A) Non-zero Elastic Net Cox model coefficients presented as hazard ratios (HR). Point color codes for marker status (blue: favorable, red: unfavorable), point size represents absolute value of HR._

_(B) Performance of the Elastic Net Cox model at prediction of biochemical relapse-free survival in the training TCGA cohort and test collectives assessed by concordance index (C-index, high values indicate high concordance between the predicted and observed survival) and integrated Brier score (IBS, low values indicate good model fit and proper calibration). Values of C-index and IBS expected for random survival prediction are depicted as dashed lines._

_(C) Fractions of biochemical relapse-free patients in low, intermediate and high tertiles of Transcriptional Collagen Score in the training TCGA cohort, and the GSE70768 and DKFZ validation collectives presented in Kaplan-Meier plots. Differences in survival between the tertiles were investigated by Peto-Peto test. Total numbers of patients and numbers of biochemical relapses are displayed in the plot captions. P values are shown in the plots._
